\citet{katz2009research} defines scientific articles as repositories of scientific observations including recipes to repeat them.
I think that is a good description of what scientific writing is about as you capture results, share how you acquired them, and then discuss the implications of your findings.
In other words, you can think of scientific writing as documenting your ideas.
Typically you will use a specific scientific method or methods to attain scientific results, and then you will capture this information in your paper in manner which often has been restricted to follow scientific format as I have discussed in the chapter \nameref{ch:elements}.

\section{Scientific results should be presented in a clean manner}

\citet{katz2009research} points out importantly that in a stereotypical paper, the results should be presented in a clean, clear, and unemotional form.
This is often one of the factors that separates scientific writing from other forms of writing as writers tend to go for some level of objectivity in their writing.
form.
That said, there has been a clear trend towards a more active form in scientific writing \citep{ryba2021better}, especially when you author sections where you presents points from your point of view, such as a part of the introduction or discussion.
Since it is a relatively new idea in scientific writing, some people still prefer passive form across their writing and you may want to check with your instructors what is the preferred way in your field.

% Another important point that \citet{katz2009research} makes is that in a stereotypical paper, the results should be presented in a clean, clear, and unemotional form.
% That said, there has been a clear trend towards a more active form in scientific writing \citep{ryba2021better}, especially when you author sections, such as a part of the introduction or discussion.

\section{Scientific writing is about transmitting your results to the reader}

\citet{gopen1990science} argue that science is often difficult to read, and likely the argument still holds today.
The main point that they make is that presentation matters and sloppy writing can obscure results and even fail to transmit vital information to the reader.
Therefore, it is our duty as scientific writers to take care of how we present our results to the reader.
Only stating the results is not enough but it should be done in a manner that is understandable to our readers that may occasionally be laymen.
Typically a good target to keep in mind is to consider that your text should be approachable at least someone in your broader field, say computer science, and include enough supporting information that they can understand what you mean.
Often this means providing enough grounding and definitions to make your text less abstract and this can help even people familiar with your field since occasionally exact definitions can differ and sometimes we use different words to mean the same things.

\section{Scientific writers translate results to a diverse set of readers}

It is difficult not to overstate the importance of considering your readers.
You will notice that people from different backgrounds and with different levels of understanding will read your work.
That is a part of the challenge of scientific writing as you will have to find ways to get your points through while living with the constraints of academic writing.
In practice, there are several ways to approach the problem and to break my personal myopia to my own content I prefer to get it read by other people quite soon after I have produced the first sensible draft.
These days tools based on artificial intelligence can provide some form of feedback but I would not rely solely on them but rather use them as a way to gain complimentary insight into your writing.
I outline several strategies I use in chapter \nameref{ch:editing} and it is with editing that you can turn a good work into great one while improving your chances of getting your papers read and into the places where you want.

% In essence, scientific writing is about transmitting your results to the reader in a way they can understand.
%The challenge is that readers come from different backgrounds and have different levels of understanding.
%Your task as a writer is to consider this problem and you act as a translator between your results and the reader.

\section{Good practices and conventions are there to help}

The lucky thing about scientific writing is that there is an established set of good practices that have formed over time that you should follow in your writing to keep it accessible to your readers.
For example, many readers expect to find a variant of so-called IMRAD structure covered in chapter \nameref{ch:elements} and following the structure gives your text a degree of readability out of the box.
There is more to scientific writing of course and you will learn more about making your text readable as you go through this book.


% Fortunately you do not have to start from scratch as there are good practices that have formed over time. For example, many readers expect to find a variant of so-called IMRAD structure covered in \nameref{ch:elements} and following the structure gives your text a degree of readability out of the box. There is more to scientific writing of course and you will learn more about making your text readable as you go through this book.

\section{How does scientific writing differ from other types of authoring?}

As mentioned above, scientific writing has a strong focus on transmitting your results to the reader.
That said, there is a lot you can do within this main constraint as a part of your work could be for example speculative as you are looking for good research problems to solve while a part of your work might include concrete measurement results.
Compared to other types of writing, you might notice that scientific literature uses so-called hedging words to give sentences a degree of uncertainty as few things are absolutely certain in science as a scientist is able to tell something about their degree of certainty through their use of words.
Another important aspect of scientific writing is that it often cannot exist in a vacuum. In other words, often you build up your writing on top of the work of others while pointing to the work of others through citations.
Citing not only shows where your ideas are coming from, it also benefits other researchers in the field by allowing them to discover related material easier.
In literature reviews, a good method is to examine references in detail to discover more papers to study to construct a good view of the research target.

\section{What are the similarities between scientific writing and other types of authoring?}

Although it might not initially feel like it, narrative plays an important role in scientific writing.
In essence you are making a case while introducing proof to support it.
From this it follows that you have to consider how present your thoughts in a logical manner while keeping the reader interested in your topic.
Another way to put it is that like any form of writing, scientific writing is about organizing your thoughts in a written manner.
Generally the more you write, the more efficient you become at this type of organizational work, and thus, writing.
It is not only writing that makes you stronger writing but also reading works of others and developing your taste.
Knowing what kind of approaches have been used before lets you improve your own style and find ways that make sense for your kind of thinking.
It is only natural that you will evolve as a writer as you produce more and your taste will become more refined.
That goes for any type of authoring as with experience nuances become more distinct and easier to appreciate.

\section{Summary}

The main task of scientific writing is to communicate your results to the reader in a clear manner.
If you can achieve that, you have already achieved a lot.
There are certain writing related specifics you should know and we will go through them in greater details in the following chapters so you become aware of good conventions and have a better idea of how to approach scientific writing.
